---
title: "R Notes for Assignment 4"
author: "Joseph S. Tabadero, Jr."
date: "March 8, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

## Preliminaries

I tried to replicate the SAS results in R. The packages that I used are:

* `survey`: for TSL estimation
* `dplyr`: for data manipulation
* `srvyr`: for using `dplyr`-like syntax that is compatible with the `survey` package.
* `ggplot2`: for graphics
* `haven`: for importing SAS data sets
* `knitr`: for printing nice tables

We can install these packages with the following command in R.

```{r eval=FALSE}
install.packages(c("survey", "dplyr", "srvyr", "ggplot2", "haven"))
```

We then load the packages using the following code.

```{r}
library(survey)
library(dplyr)
library(srvyr)
library(ggplot2)
library(haven)
```


## Problem 1

The following code imports the data into R.

```{r}
library(haven)
# import the data and assign it to the object ncsr
ncsr <- read_sas("~/Stat_291/ncsr.sas7bdat")
```

To view the data set `ncsr`, we can type:

```{r eval = FALSE}
View(ncsr)
```

Now we create an additional column `NCSRWTSH_POP` as required by problem 1 in the problem set.

```{r}
ncsr <- ncsr %>% 
  mutate(NCSRWTSH_POP = 209128094/9282*NCSRWTSH)
```


The following data creates a survey design object in R. 

```{r}
# name the design object ncsr_dsgn
ncsr_dsgn <- ncsr %>%
# specify the cluster (ids), strata, weight, and set 
# tell R that clusters are nested inside strata
  as_survey_design(ids = SECLUSTR, 
                   strata = SESTRAT,
                   weight = NCSRWTSH_POP,
                   nest = TRUE)
```


Firstly, I want to print the number of strata, clusters, observations, and sum of weights in R. This is not very straightforward as the `PROC SURVEYMEANS` or `PROC SURVEYFREQ` in SAS. (I might write the package developers to include these features in the future.)

```{r}
ncsr_dsgn %>%
  summarise(
    `Number of strata` = count
  )
```

