---
title: "How to conduct a t test in R?"
author: "Joseph S. Tabadero, Jr."
date: "January 20, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Requirements

* One independent, categorical variable that has two levels/groups.
* One continuous, dependent variable

## Assumptions

* Data collected follows a continuous or ordinal scale
* Data is collected through randomly assigning the two levels of the independet variable
* Unrelated, unpaired, independent groups (levels of the independent variable)
* The errors are normally distributed
* Homogeneity of variance

```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
```


## The data

```{r}
data("mtcars")
glimpse(mtcars)
mtcars <- mtcars %>% mutate(am = as.factor(am))
with(mtcars,t.test(mpg ~ am, var.equal=TRUE))
t.test(mtcars$mpg ~ mtcars$am, var.equal = TRUE)
with(mtcars, t.test(mpg ~ am))
library(forcats)
data <- mtcars %>% mutate(am = fct_recode(am, "auto"="0", "manual"="1"))
data <- data %>% spread(am, mpg)
t.test(data$auto, data$manual, var.equal = TRUE)
```
```{r}
ggplot(mtcars, aes(am, mpg)) + geom_boxplot()
```

```{r}
ggplot(mtcars, aes(mpg)) + geom_histogram(bins=5) + facet_wrap(~am)
```

```{r}
auto = mtcars$mpg[mtcars$am=="0"]
manual = mtcars$mpg[mtcars$am=="1"]
manual = mtcars %>% filter(am==1) %>% select(mpg)
manual

model <- lm(mpg ~ am, data = mtcars)
qqnorm(residuals(model), ylab="Residuals")
qqline(residuals(model))
shapiro.test(manual)
auto
```

